BayesPy: Variational Bayesian Inference in Python

### 2015-06-05 bayespy.tex
Revised version

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
  BayesPy is an open-source Python software package for performing variational
  Bayesian inference.  It is based on the variational message passing framework
  and supports conjugate exponential family models.  By removing the tedious
  task of implementing the variational Bayesian update equations, the user can
  construct models faster and in a less error-prone way.  Simple syntax,
  flexible model construction and efficient inference make BayesPy suitable for
  both average and expert Bayesian users.
\end{abstract}

@@

BayesPy is a Python 3 package providing tools for constructing conjugate
exponential family models and performing VB inference easily and efficiently.
It is based on the variational message passing (VMP) framework which defines a
simple message passing protocol \citep{Winn:2005}.  This enables implementation
of small general nodes that can be used as building blocks for large complex
models.  BayesPy offers a comprehensive collection of built-in nodes that can be
used to build a wide range of models and a simple interface for implementing
custom nodes.  The package is released under the GNU General Public License v3.0
(GPLv3).

@@

BayesPy can be used to construct conjugate exponential family models.  The
documentation provides detailed examples of how to construct a variety of
models, including principal component analysis models, linear state-space
models, mixture models and hidden Markov models.  In addition, BayesPy has been
used in two publications about parameter expansion and time-varying dynamics for
linear state-space models \citep{Luttinen:2013,Luttinen:2014}.
% , which show that it is applicable to large complex models requiring efficient
% inference.

@@

This section demonstrates the key steps in using BayesPy.  An artificial
Gaussian mixture dataset is created by drawing 500 samples from two
2-dimensional Gaussian distributions.  200 samples have mean $[2, 2]$ and 300
samples have mean $[0, 0]$:
\begin{lstlisting}
import numpy as np
N = 500; D = 2
data = np.random.randn(N, D)
data[:200,:] += 2*np.ones(D)
\end{lstlisting}
We construct a mixture model for the data and assume that the parameters, the
cluster assignments and the true number of clusters are unknown.  The model uses
a maximum number of five clusters but the effective number of clusters will be
determined automatically:
\begin{lstlisting}
K = 5
\end{lstlisting}
The unknown cluster means and precision matrices are given Gaussian and Wishart
prior distributions:
\begin{lstlisting}
from bayespy import nodes
mu = nodes.Gaussian(np.zeros(D), 0.01*np.identity(D), plates=(K,))
Lambda = nodes.Wishart(D, D*np.identity(D), plates=(K,))
\end{lstlisting}
The \texttt{plates} keyword argument is used to define repetitions similarly as
the plate notation in graphical models.  The cluster assignments are categorical
variables, and the cluster probabilities are given a Dirichlet prior
distribution:
\begin{lstlisting}
alpha = nodes.Dirichlet(0.01*np.ones(K))
z = nodes.Categorical(alpha, plates=(N,))
\end{lstlisting}
The observations are from a Gaussian mixture distribution:
\begin{lstlisting}
y = nodes.Mixture(z, nodes.Gaussian, mu, Lambda)
\end{lstlisting}
The second argument for the \texttt{Mixture} node defines the type of the
mixture distribution, in this case Gaussian.  The variable is marked as observed
by providing the data:
\begin{lstlisting}
y.observe(data)
\end{lstlisting}
Next, we want to find the posterior approximation for our latent variables.  We
create the variational Bayesian inference engine:
\begin{lstlisting}
from bayespy.inference import VB
Q = VB(y, mu, z, Lambda, alpha)
\end{lstlisting}
Before running the VMP algorithm, the symmetry in the model is broken by
a random initialization of the cluster assignments:
\begin{lstlisting}
z.initialize_from_random()
\end{lstlisting}
Without the random initialization, the clusters would not be separated.  The VMP
algorithm updates the variables in turns and is run for 200 iterations or until
convergence:
\begin{lstlisting}
Q.update(repeat=200)
\end{lstlisting}
The results can be examined visually by using \texttt{bayespy.plot} module:
\begin{lstlisting}
import bayespy.plot as bpplt
bpplt.gaussian_mixture_2d(y, alpha=alpha)
bpplt.pyplot.show()
\end{lstlisting}
It is also possible to print the parameters of the approximate posterior
distributions:
\begin{lstlisting}
print(alpha)
\end{lstlisting}
The \texttt{bayespy.plot} module contains also other functions for visually
examining the distributions.
%In addition, module \texttt{bayespy.plot} contains several plotting functions to
%examine the results visually.

@@

The speed of the packages were compared by using two widely used models: a
Gaussian mixture model (GMM) and principal component analysis
(PCA).\footnote{The scripts for running the experiments are available as
  supplementary material.}  Both models were run for small and large
artificially generated datasets.  For GMM, the small model used 10 clusters for
200 observations with 2 dimensions, and the large model used 40 clusters for
2000 observations with 10 dimensions.  For PCA, the small model used
10-dimensional latent space for 500 observations with 20 dimensions, and the
large model used 40-dimensional latent space for 2000 observations with 100
dimensions.
For the PCA model, Infer.NET used a fully factorizing approximation and also
similarly to BayesPy an approximation which did not factorize with respect to
the latent space dimensions.
%Infer.NET used a fully factorizing approximation for the PCA model,
%whereas BayesPy did not factorize with respect to the latent space dimensions.  
The experiments were run on a Linux computer with a quad-core processor
(i7-4702MQ).  Because the packages run practically identical algorithms and thus
converged to similar solutions in a similar number of iterations (50--100
iterations depending on the dataset), we compared the average CPU time per
iteration.

The results are summarized in Table~\ref{tab:speed}.  For GMM, BayesPy is
significantly faster, which is probably mostly explained by the fact that
BayesPy uses highly optimized numerical libraries (BLAS and LAPACK).  For PCA,
BayesPy automatically avoids redundant computations which arise because equal
posterior covariance matrices are shared between latent variables.  This
broadcasting makes BayesPy significantly faster than Infer.NET.  However,
broadcasting is not always possible (e.g., if there are missing values in the
PCA data), thus the table also presents the results when BayesPy is forced not
to utilize broadcasting.  In that case, BayesPy is a few times slower than
Infer.NET because Infer.NET uses a fully factorized approximation.  If Infer.NET
uses similar factorization as BayesPy, Infer.NET may be orders of magnitude
slower.
% On the smaller datasets,
% Infer.NET is slightly faster because BayesPy has some Python overhead.  However,
% the total CPU time for both methods is only a few seconds.  On the larger
% datasets, BayesPy is significantly faster because it uses BLAS and LAPACK
% libraries for fast numerical operations.\footnote{Assuming that NumPy uses
%   BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}  In
% addition, if using multi-threaded BLAS and LAPACK on a multi-core computer, the
% difference in wall-clock time can be even larger than in CPU time.
The speed of BayesPy for both models might be improved by using collapsed or
stochastic variational inference.  The accuracy of BayesPy could be improved by
using a Gaussian-Wishart node, which does not factorize with respect to the
Gaussian and Wishart variables.

\begin{table}[tb]
  \centering
  \caption{
    The average CPU time (in milliseconds) per iteration for each
    dataset.  The results in parentheses have the following meanings: a) BayesPy without utilizing
    broadcasting. b) Infer.NET using an approximation factorized similarly as in
    BayesPy.
  }
  \small
  \begin{tabular}{ccccc}
    &
    Small GMM
    &
    Large GMM
    &
    Small PCA
    &
    Large PCA
    \\
    \hline
    BayesPy     & 5 & 85  & 7 (74) & 360 (6\,200)
    \\
    Infer.NET   & 25 & 4\,600 & 37 (350) & 2\,200 (210\,000)
  \end{tabular}
  \label{tab:speed}
\end{table}

>>>>

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
  BayesPy is an open-source Python software package for performing variational
  Bayesian inference.  It is based on the variational message passing framework
  and supports conjugate exponential family models.  By removing the tedious
  task of implementing the variational Bayesian update equations, the user can
  construct models faster and in a less error-prone way.  Simple syntax,
  flexible model construction and efficient inference make BayesPy suitable for
  both average and expert Bayesian users.  It also supports some advanced
  methods such as stochastic and collapsed variational inference.
\end{abstract}

@@

BayesPy is a Python 3 package providing tools for constructing conjugate
exponential family models and performing VB inference easily and efficiently.
It is based on the variational message passing (VMP) framework which defines a
simple message passing protocol \citep{Winn:2005}.  This enables implementation
of small general nodes that can be used as building blocks for large complex
models.  BayesPy offers a comprehensive collection of built-in nodes that can be
used to build a wide range of models and a simple interface for implementing
custom nodes.  The package is released under the MIT license.

@@

BayesPy can be used to construct conjugate exponential family models.  The
documentation provides detailed examples of how to construct a variety of
models, including principal component analysis models, linear state-space
models, mixture models and hidden Markov models.  BayesPy has also been used in
two publications about parameter expansion and time-varying dynamics for linear
state-space models \citep{Luttinen:2013,Luttinen:2014}.
% , which show that it is applicable to large complex models requiring efficient
% inference.

@@

This section demonstrates the key steps in using BayesPy.  An artificial
Gaussian mixture dataset is created by drawing 500 samples from two
2-dimensional Gaussian distributions.  200 samples have mean $[2, 2]$ and 300
samples have mean $[0, 0]$:
\begin{lstlisting}
import numpy as np
N = 500; D = 2
data = np.random.randn(N, D)
data[:200,:] += 2*np.ones(D)
\end{lstlisting}
We construct a mixture model for the data and assume that the parameters, the
cluster assignments and the true number of clusters are unknown.  The model uses
a maximum number of five clusters but the effective number of clusters will be
determined automatically:
\begin{lstlisting}
K = 5
\end{lstlisting}
The unknown cluster means and precision matrices are given Gaussian and Wishart
prior distributions:
\begin{lstlisting}
from bayespy import nodes
mu = nodes.Gaussian(np.zeros(D), 0.01*np.identity(D), plates=(K,))
Lambda = nodes.Wishart(D, D*np.identity(D), plates=(K,))
\end{lstlisting}
The \texttt{plates} keyword argument is used to define repetitions similarly to
the plate notation in graphical models.  The cluster assignments are categorical
variables, and the cluster probabilities are given a Dirichlet prior
distribution:
\begin{lstlisting}
alpha = nodes.Dirichlet(0.01*np.ones(K))
z = nodes.Categorical(alpha, plates=(N,))
\end{lstlisting}
The observations are from a Gaussian mixture distribution:
\begin{lstlisting}
y = nodes.Mixture(z, nodes.Gaussian, mu, Lambda)
\end{lstlisting}
The second argument for the \texttt{Mixture} node defines the type of the
mixture distribution, in this case Gaussian.  The variable is marked as observed
by providing the data:
\begin{lstlisting}
y.observe(data)
\end{lstlisting}
Next, we want to find the posterior approximation for our latent variables.  We
create the variational Bayesian inference engine:
\begin{lstlisting}
from bayespy.inference import VB
Q = VB(y, mu, z, Lambda, alpha)
\end{lstlisting}
Before running the VMP algorithm, the symmetry in the model is broken by
a random initialization of the cluster assignments:
\begin{lstlisting}
z.initialize_from_random()
\end{lstlisting}
Without the random initialization, the clusters would not be separated.  The VMP
algorithm updates the variables in turns and is run for 200 iterations or until
convergence:
\begin{lstlisting}
Q.update(repeat=200)
\end{lstlisting}
The results can be examined visually by using \texttt{bayespy.plot} module:
\begin{lstlisting}
import bayespy.plot as bpplt
bpplt.gaussian_mixture_2d(y, alpha=alpha)
bpplt.pyplot.show()
\end{lstlisting}
It is also possible to print the parameters of the approximate posterior
distributions:
\begin{lstlisting}
print(alpha)
\end{lstlisting}
The \texttt{bayespy.plot} module contains also other functions for visually
examining the distributions.
%In addition, module \texttt{bayespy.plot} contains several plotting functions to
%examine the results visually.

@@

The speed of the packages were compared by using two widely used models: a
Gaussian mixture model (GMM) and principal component analysis
(PCA).\footnote{The scripts for running the experiments are available as
  supplementary material.}  Both models were run for small and large artificial
datasets.  For GMM, the small model used 10 clusters for 200 observations with 2
dimensions, and the large model used 40 clusters for 2000 observations with 10
dimensions.  For PCA, the small model used 10-dimensional latent space for 500
observations with 20 dimensions, and the large model used 40-dimensional latent
space for 2000 observations with 100 dimensions.  For the PCA model, Infer.NET
used both a fully factorizing approximation and also the same approximation as
in BayesPy which did not factorize with respect to the latent space dimensions.
%Infer.NET used a fully factorizing approximation for the PCA model,
%whereas BayesPy did not factorize with respect to the latent space dimensions.  
The experiments were run on a quad-core (i7-4702MQ) Linux computer.
% with a quad-core processor (i7-4702MQ).
Because the packages run practically identical algorithms and thus
converged to similar solutions in a similar number of iterations (50--100
iterations depending on the dataset), we compared the average CPU time per
iteration.

The results are summarized in Table~\ref{tab:speed}.  For all datasets, BayesPy
is faster probably because it uses highly optimized numerical libraries (BLAS
and LAPACK).  For PCA, BayesPy also automatically avoids redundant computations
which arise because latent variables have equal posterior covariance matrices.
% This broadcasting makes BayesPy significantly faster than Infer.NET.
However, such broadcasting is not always possible (e.g., if there are missing
values in the PCA data).  Thus, the table also presents the results when BayesPy
is forced to not use broadcasting: BayesPy is slower than Infer.NET on the
smaller PCA dataset but faster on the larger PCA dataset.
% Note that Infer.NET uses a fully factorized approximation.
If Infer.NET used the same factorization for PCA as BayesPy, Infer.NET may be
orders of magnitude slower.
% On the smaller datasets,
% Infer.NET is slightly faster because BayesPy has some Python overhead.  However,
% the total CPU time for both methods is only a few seconds.  On the larger
% datasets, BayesPy is significantly faster because it uses BLAS and LAPACK
% libraries for fast numerical operations.\footnote{Assuming that NumPy uses
%   BLAS/LAPACK because BayesPy uses NumPy for numerical operations.}  In
% addition, if using multi-threaded BLAS and LAPACK on a multi-core computer, the
% difference in wall-clock time can be even larger than in CPU time.
%
%The speed of BayesPy for both models might be improved by using collapsed or
%stochastic variational inference.  The accuracy of BayesPy could be improved by
%using a Gaussian-Wishart node, which does not factorize with respect to the
%Gaussian and Wishart variables.

\begin{table}[tb]
  \centering
  \caption{
    The average CPU time in milliseconds per iteration for each
    dataset.  The results in parentheses have the following meanings: a) BayesPy without using
    broadcasting. b) Infer.NET using the same factorization as in
    BayesPy.
  }
  \small
  \begin{tabular}{ccccc}
    &
    Small GMM
    &
    Large GMM
    &
    Small PCA
    &
    Large PCA
    \\
    \hline
    BayesPy     & 6 & 90  & 7 (60) & 400 (1\,500)
    \\
    Infer.NET   & 25 & 4\,600 & 37 (350) & 2\,200 (210\,000)
  \end{tabular}
  \label{tab:speed}
\end{table}
